---
title: "Galaxy Stats"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: 
      collapsed: true
    toc_depth: 4
---

# Setup

```{r message=FALSE}
# install.packages("RPostgreSQL")
# install.packages("lubridate")
# install.packages("openxlsx")
# install.packages("purrr")
library(RPostgreSQL)
library(dplyr)
library(stringr)
library(lubridate)
library(purrr)
library(ggplot2)
library(openxlsx)
```

```{r}
year_month <- function(date) {
  return(sprintf("%d-%02d", year(date), month(date)))
}
```

Get current year and month.

```{r}
current_year_month <- year_month(Sys.Date())
previous_year_month <- year_month(Sys.Date() - months(1))
```

Load driver and create a connection.

```{r}
drv <- dbDriver("PostgreSQL")
con <- dbConnect(drv, dbname = "mel-backup", host = "localhost", port = 5950,
                 user = "galaxy")
````

## Get data

Get and store user data.

```{r}
# dbGetQuery(con, "SELECT * FROM galaxy_user LIMIT 5;")
all_users <- dbGetQuery(con, "SELECT * FROM galaxy_user;") %>% 
  select(id:email, username, disk_usage)
```

Get and store job data.

```{r}
# dbGetQuery(con, "SELECT * FROM job LIMIT 5;")
cmd <- paste0("SELECT id, create_time, history_id, tool_id, ",
              "tool_version, state, command_line FROM job;")
all_jobs <- dbGetQuery(con, cmd)
```

Get and store dataset data.

```{r}
# dbGetQuery(con, "SELECT * FROM dataset LIMIT 5;")
cmd <- paste0("SELECT id, create_time, total_size FROM dataset;")
all_datasets <- dbGetQuery(con, cmd)
```

Get and store workflow data.

```{r}
# dbGetQuery(con, "SELECT * FROM stored_workflow LIMIT 5;")
cmd <- paste0("SELECT id, create_time FROM stored_workflow;")
all_workflows <- dbGetQuery(con, cmd)
```


```{r include=FALSE, eval=FALSE}
# How much memory is being used to store the dataframes?
# Not much. Less than 20 MB per data frame.
sapply(c("all_users", "all_jobs", "all_datasets", "all_workflows"), function(x){
  object.size(get(x)) %>% format(units = "MB")})
```

-----

# Analysis

## User analysis

New users per month.

```{r}
# Create new column with year-month
users <- all_users %>% 
  mutate(month=year_month(create_time))

# Get number of new users per year-month
users_by_month <- users %>%
  group_by(month) %>% 
  summarize(n=n())

users_by_month %>% tail(12)
```

```{r include=FALSE}
# ggplot(users, aes(x=create_time %>% date)) +
#   geom_histogram(bins=20) +
#   scale_x_date(date_breaks="3 months") +
#   labs(x="date", y="new users")
# 
# ggplot(users, aes(x=create_time %>% ymd_hms)) +
#   geom_density()
```

```{r}
ggplot(users, aes(x=month)) + 
  geom_bar(alpha=0.8) +
  labs(x="date", y="new users") +
  theme(axis.text.x=element_text(angle=90, hjust=1))
```

#### Email domains

Analysis of email suffixes.

```{r}
# Most common domains
email_suffixes <- users$email %>% str_split("@") %>% sapply(function(x) x[2])
email_suffix_df <- email_suffixes %>% table %>% 
  data.frame %>% setNames(c("domain", "freq")) %>%
  arrange(desc(freq))
email_suffix_df %>% head(10)
```

```{r}
# Country codes
country_codes <- email_suffixes %>% str_split("\\.") %>% 
  sapply(function(x) {x[length(x)]})
country_code_df <- country_codes %>% table %>% 
  data.frame %>% setNames(c("country_codes", "freq")) %>%
  arrange(desc(freq))
country_code_df %>% head(10)
```

```{r}
# edu domains
edu_addresses <- email_suffixes %>% str_split("\\.") %>% 
  sapply(function(x) {"edu" %in% x}) %>% email_suffixes[.]
edu_address_df <- edu_addresses %>% table %>% 
  data.frame %>% setNames(c("domain", "freq")) %>%
  arrange(desc(freq))
edu_address_df %>% head(10)
```

## Job analysis

Remove jobs that occurred before the first user creation (Not sure why some jobs have incorrect timestamps).

```{r}
earliest_user <- users$create_time %>% sort %>% .[1]
earliest_user
```

```{r}
# Check if date is before earliest user
all_jobs <- all_jobs %>% mutate(ok_date=create_time > earliest_user)

# Filter out bad dates
jobs <- all_jobs %>% filter(ok_date)
not_ok <- all_jobs %>% filter(! ok_date)

# How many dates are ok vs not ok?
all_jobs$ok_date %>% table
```

```{r include=FALSE, eval=FALSE}
# Using Galaxy-tut database:
# Why are there dates that are before the earliest created user?
# Are these jobs uniformly distributed or is there a pattern?
ggplot(all_jobs, aes(x=id, y=ok_date)) + geom_point(size=0.1)
# Looks like they're clusted in some areas

# The jobs with incorrect dates seem to be from the same 4 tools
not_ok$tool_id %>% table

# But only sometimes for filtering and uploading, and always for bwa and
# freebayes
tool_subset <- not_ok$tool_id %>% unique
all_jobs %>% filter(tool_id %in% tool_subset) %>% group_by(tool_id) %>%
  summarize(n=n(), ok=sum(ok_date), not_ok=sum(!ok_date))

# Rename tool_ids of bwa and freebayes
not_ok <- not_ok %>% mutate(
  tool_id=ifelse(tool_id == "toolshed.g2.bx.psu.edu/repos/devteam/bwa/bwa/0.3.1",
                 "bwa-0.3.1", tool_id),
  tool_id=ifelse(tool_id == "toolshed.g2.bx.psu.edu/repos/devteam/freebayes/freebayes/0.3",
                 "freebayes-0.3", tool_id))
  
# Is there a tool order pattern? Is it one particular tool which triggers it?
ggplot(not_ok %>% head(100), aes(x=id, y=tool_id)) +
  geom_point(size=0.1)

# Looks like bwa -> freebayes -> filter is the culprit for most of them
not_ok %>% arrange(id) %>% select(id:tool_id) %>% head(10)
```

Jobs run per month.

```{r}
# Create new column with year-month
jobs <- jobs %>% 
  mutate(month=year_month(create_time))

# Get number of jobs run per year-month
jobs_by_month <- jobs %>%
  group_by(month) %>% 
  summarize(n=n())

jobs_by_month %>% tail(12)
```

```{r include=FALSE}
# ggplot(jobs, aes(x=create_time %>% date)) +
#   geom_histogram(bins=20) +
#   scale_x_date(date_breaks="3 months") +
#   labs(x="date", y="jobs")
# 
# ggplot(jobs, aes(x=create_time %>% ymd_hms)) +
#   geom_density()
```

```{r}
ggplot(jobs, aes(x=month)) + 
  geom_bar(alpha=0.8) +
  labs(x="date", y="number of jobs") +
  theme(axis.text.x=element_text(angle=90, hjust=1))
```

## Tool analysis

Which tools are used most often?

```{r}
# Using all jobs across all dates
tool_ids <- jobs$tool_id
tool_data <- tool_ids %>% str_split("/")
tool_names <- sapply(tool_data, function(x) {ifelse(length(x) == 1, x, x[5])})

# Top 20 used tools
top_tool_df <- tool_names %>% table %>% data.frame %>% 
  setNames(c("tool", "freq")) %>% arrange(desc(freq))
top_tool_df %>% head(20)
```

## Reference analysis

Which reference genomes are frequently used?

For DNA alignment?

```{r}
# From DNA alignment tools (bwa? bowtie?)
dna_jobs <- jobs %>% mutate(tool_name=tool_names) %>% 
  filter(str_detect(tool_name, "bwa|bowtie"))
dna_ref_df <- dna_jobs$command %>% 
  str_extract("/mnt/galaxyIndices/[A-Za-z0-9/.-_]+") %>%
  str_split("/") %>% sapply(function(x) x[length(x)]) %>%
  str_replace(".fa$", "") %>%
  table() %>% data.frame() %>% setNames(c("reference", "freq")) %>%
  arrange(desc(freq))
dna_ref_df
```

For RNA alignment?

```{r}
# From RNA alignment tools (tophat? hisat?)
rna_jobs <- jobs %>% mutate(tool_name=tool_names) %>% 
  filter(str_detect(tool_name, "tophat|hisat"))
rna_ref_df <- rna_jobs$command %>% 
  str_extract("/mnt/galaxyIndices/[A-Za-z0-9/.-_]+") %>%
  str_split("/") %>% sapply(function(x) x[length(x)]) %>%
  str_replace(".fa$", "") %>%
  table() %>% data.frame() %>% setNames(c("reference", "freq")) %>%
  arrange(desc(freq))
rna_ref_df
```

Across all jobs?

```{r}
# Get only the first instance of the string match
all_ref_df <- jobs$command %>%
  str_extract("/mnt/galaxyIndices/genomes/[A-Za-z0-9]+") %>%
  .[!is.na(.)] %>%
  str_replace("/mnt/galaxyIndices/genomes/","") %>% 
  table() %>% data.frame() %>% setNames(c("reference", "freq")) %>%
  arrange(desc(freq))
all_ref_df
```

## Dataset analysis

How many new datasets created per month?

```{r}
# Create new column with year-month
datasets <- all_datasets %>% 
  mutate(month=year_month(create_time))

# Get number of datasets run per year-month
datasets_by_month <- datasets %>%
  group_by(month) %>% 
  summarize(n=n())

datasets_by_month %>% tail(12)
```

```{r}
ggplot(datasets, aes(x=month)) + 
  geom_bar(alpha=0.8) +
  labs(x="date", y="new datasets") +
  theme(axis.text.x=element_text(angle=90, hjust=1))
```

Total size of newly created datasets per month.

```{r}
dataset_sizes_by_month <- datasets %>%
  group_by(month) %>% 
  summarize(size_in_bytes=sum(total_size, na.rm=TRUE)) %>%
  mutate(size_in_gb=round(size_in_bytes/(2^30), digits=0))

dataset_sizes_by_month %>% tail(12)
```

## Workflow analysis

How many new workflows created per month?

```{r}
# Create new column with year-month
workflows <- all_workflows %>% 
  mutate(month=year_month(create_time))

# Get number of workflows created per year-month
workflows_by_month <- workflows %>%
  group_by(month) %>% 
  summarize(n=n())

workflows_by_month %>% tail(12)
```

```{r}
ggplot(workflows, aes(x=month)) + 
  geom_bar(alpha=0.8) +
  labs(x="date", y="new workflows") +
  theme(axis.text.x=element_text(angle=90, hjust=1))
```

-----

# Files for RDS

Get data for the RDS Genomics dashboard in the format described in the emails.

## month_stat-item_stat-data

Columns: month, stat_item, stat_data, description  
Rows: workflow, job, datasets_created, datasets_created_storage

```{r}
df1 <- workflows_by_month %>%
  mutate(stat_item="workflow",
         stat_data=n,
         description=map2_chr(stat_item, month, function(x, y) {
           sprintf("Number of %ss created for month %s", x, y)
         })) %>%
  select(month, stat_item, stat_data, description)

df2 <- jobs_by_month %>%
  mutate(stat_item="job",
         stat_data=n,
         description=map2_chr(stat_item, month, function(x, y) {
           sprintf("Number of %ss executed for month %s", x, y)
         })) %>%
  select(month, stat_item, stat_data, description)

df3 <- datasets_by_month %>%
  mutate(stat_item="datasets_created",
         stat_data=n,
         description=map2_chr(stat_item, month, function(x, y) {
           sprintf("Number of %s for month %s", x, y)
         })) %>%
  select(month, stat_item, stat_data, description)

df4 <- dataset_sizes_by_month %>%
  mutate(stat_item="datasets_created_storage",
         stat_data=size_in_gb,
         description=map2_chr(stat_item, month, function(x, y) {
           sprintf("On-disk size of %s in gigabytes for month %s", x, y)
         })) %>%
  select(month, stat_item, stat_data, description)

stat_output <- rbind(df1, df2, df3, df4) %>%
  filter(month != current_year_month) %>%
  arrange(desc(month))
```

## month_email-domain_count

Columns: month, email_domain, count, description  
Description: "Number of users with the email_domain for month YYYY-MM"

```{r}
email_suffix_by_month <- users %>%
  mutate(suffix=email %>% str_split("@") %>% sapply(function(x) x[2])) %>%
  group_by(month, suffix) %>%
  summarise(n=n()) %>% 
  filter(month != current_year_month) %>%
  arrange(month, suffix)
```


```{r}
email_output <- email_suffix_by_month %>% 
  mutate(email_domain=suffix,
         count=n,
         description=map_chr(month, function(x) {
           paste("Number of new users with the email_domain for", x)
         })) %>%
  select(month, email_domain, count, description)
```

## month_tool_count

Columns: month, tool, count, description  
Description: "Number of tool usage for month YYYY-MM"

```{r}
stopifnot(length(jobs$month) == length(tool_names))
tool_by_month <- data.frame(month=jobs$month, tool=tool_names) %>%
  group_by(month, tool) %>%
  summarise(n=n()) %>%
  filter(month != current_year_month) %>%
  arrange(month, tool)
```

```{r}
tool_output <- tool_by_month %>%
  mutate(count=n,
         description=map_chr(month, function(x) {
           paste("Number of instances of tool usage for the month of", x)
         })) %>%
  select(month, tool, count, description)
```

## month_reference-genome_count

Columns: month, reference_genome, count, description  
Description: "Number of times the reference_genome is used for month YYYY-MM"

```{r}
ref_by_month <- jobs %>% 
  mutate(genome=command_line %>% 
           str_extract("/mnt/galaxyIndices/genomes/[A-Za-z0-9]+")) %>%
  filter(! is.na(genome)) %>%
  mutate(genome=str_replace(genome, "/mnt/galaxyIndices/genomes/", "")) %>% 
  select(month, genome) %>%
  group_by(month, genome) %>%
  summarise(n=n()) %>%
  arrange(month, genome)
```


```{r}
ref_output <- ref_by_month %>%
  mutate(reference_genome=genome,
         count=n,
         description=map_chr(month, function(x) {
           paste("Number of times the reference_genome was used for",
                 "the month of", x)
         })) %>%
  select(month, reference_genome, count, description)
```

-----

# Other output files

```{r}
# Merge users, jobs, datasets, workflows per month to a single dataframe
monthly_stats <- 
  merge(users_by_month %>% rename(new_users=n),
        jobs_by_month %>% rename(new_jobs=n),
        all=TRUE, by="month") %>%
  merge(datasets_by_month %>% rename(new_datasets=n),
        all=TRUE, by="month") %>%
  merge(dataset_sizes_by_month %>% select(-size_in_bytes) %>%
          rename(new_datasets_gb=size_in_gb),
        all=TRUE, by="month") %>%
  merge(workflows_by_month %>% rename(new_workflows=n),
        all=TRUE, by="month") %>%
  arrange(month)
```

```{r include=FALSE, eval=FALSE}
# May have some NA values for zero counts
monthly_stats
```

```{r}
# Replace any NA values with zero
monthly_stats[is.na(monthly_stats)] <- 0
```

-----

# Write to files

## RDS files

```{r}
write.table(stat_output, file="month_stat-item_stat-data.tsv",
            quote=FALSE, sep="\t", row.names=FALSE)
write.table(email_output, file="month_email-domain_count.tsv",
            quote=FALSE, sep="\t", row.names=FALSE)
write.table(tool_output, file="month_tool_count.tsv",
            quote=FALSE, sep="\t", row.names=FALSE)
write.table(ref_output, file="month_reference-genome_count.tsv",
            quote=FALSE, sep="\t", row.names=FALSE)
```

## Other files

```{r}
write.table(monthly_stats, file="monthly_stats.tsv",
            quote=FALSE, sep="\t", row.names=FALSE)
```

```{r}
output_list <- list("Monthly stats" = monthly_stats,
                    "Email - suffixes" = email_suffix_df,
                    "Email - countries" = country_code_df,
                    "Email - edu addresses" = edu_address_df,
                    "Tools" = top_tool_df,
                    "Reference - DNA" = dna_ref_df,
                    "Reference - RNA" = rna_ref_df
                    )
write.xlsx(output_list, file="galaxy_stats.xlsx")
```

-----

# Session Info

```{r}
devtools::session_info()
```

