---
title: "Galaxy Stats"
output: html_document
---

```{r}
# install.packages("RPostgreSQL")
# install.packages("lubridate")
library("RPostgreSQL")
library("dplyr")
library("stringr")
library("lubridate")
library("ggplot2")
```

```{r}
year_month <- function(date) {
  return(sprintf("%d-%02d", year(date), month(date)))
}
```

Load driver and create a connection.

```{r}
drv <- dbDriver("PostgreSQL")
con <- dbConnect(drv, dbname = "galaxy", host = "localhost", port = 5930,
                 user = "ubuntu")
````

## Get data

Get and store user data.

```{r}
# dbGetQuery(con, "SELECT * FROM galaxy_user LIMIT 5;")
users <- dbGetQuery(con, "SELECT * FROM galaxy_user;") %>% 
  select(id:email, username, disk_usage)
```

Get and store job data.

```{r}
# dbGetQuery(con, "SELECT * FROM job LIMIT 5;")
cmd <- paste0("SELECT id, create_time, update_time, history_id, tool_id, ",
              "tool_version, state, command_line FROM job;")
all_jobs <- dbGetQuery(con, cmd)
```

## User analysis

New users per month.

```{r}
# Create new column with year-month
users <- users %>% 
  mutate(month=year_month(create_time))

# Get number of new users per year-month
users %>%
  group_by(month) %>% 
  summarize(n=n())
```

```{r include=FALSE}
# ggplot(users, aes(x=create_time %>% date)) +
#   geom_histogram(bins=20) +
#   scale_x_date(date_breaks="3 months") +
#   labs(x="date", y="new users")
# 
# ggplot(users, aes(x=create_time %>% ymd_hms)) +
#   geom_density()
```

```{r}
ggplot(users, aes(x=month)) + 
  geom_bar(alpha=0.8) +
  labs(x="date", y="new users")
```

Analysis of email suffixes.

```{r}
# Most common domains
email_suffixes <- users$email %>% str_split("@") %>% sapply(function(x) x[2])
email_suffixes %>% table %>% 
  data.frame %>% setNames(c("domain", "freq")) %>%
  arrange(desc(freq)) %>% head(20)
```

```{r}
# Country codes
country_codes <- email_suffixes %>% str_split("\\.") %>% 
  sapply(function(x) {x[length(x)]})
country_codes %>% table %>% 
  data.frame %>% setNames(c("country_codes", "freq")) %>%
  arrange(desc(freq)) %>% head(10)
```

```{r}
# edu domains
edu_addresses <- email_suffixes %>% str_split("\\.") %>% 
  sapply(function(x) {"edu" %in% x}) %>% email_suffixes[.]
edu_addresses %>% table %>% 
  data.frame %>% setNames(c("domain", "freq")) %>%
  arrange(desc(freq)) %>% head(20)
```

## Job analysis

Remove jobs that occurred before the first user creation (Not sure why some jobs have incorrect timestamps).

```{r}
earliest_user <- users$create_time %>% sort %>% .[1]
earliest_user
```

```{r}
# Check if date is before earliest user
all_jobs <- all_jobs %>% mutate(ok_date=create_time > earliest_user)

# Filter out bad dates
jobs <- all_jobs %>% filter(ok_date)

# How many dates are ok vs not ok?
all_jobs$ok_date %>% table
```

```{r include=FALSE, eval=FALSE}
# Why are there dates that are before the earliest created user?
# Are these jobs uniformly distributed or is there a pattern?
ggplot(all_jobs, aes(x=id, y=ok_date)) + geom_point(size=0.1)
# Looks like they're clusted in some areas

# The jobs with incorrect dates seem to be from the same 4 tools
not_ok$tool_id %>% table

# But only sometimes for filtering and uploading, and always for bwa and
# freebayes
tool_subset <- not_ok$tool_id %>% unique
all_jobs %>% filter(tool_id %in% tool_subset) %>% group_by(tool_id) %>%
  summarize(n=n(), ok=sum(ok_date), not_ok=sum(!ok_date))

# Rename tool_ids of bwa and freebayes
not_ok <- not_ok %>% mutate(
  tool_id=ifelse(tool_id == "toolshed.g2.bx.psu.edu/repos/devteam/bwa/bwa/0.3.1",
                 "bwa-0.3.1", tool_id),
  tool_id=ifelse(tool_id == "toolshed.g2.bx.psu.edu/repos/devteam/freebayes/freebayes/0.3",
                 "freebayes-0.3", tool_id))
  
# Is there a tool order pattern? Is it one particular tool which triggers it?
ggplot(not_ok %>% head(100), aes(x=id, y=tool_id)) +
  geom_point(size=0.1)

# Looks like bwa -> freebayes -> filter is the culprit for most of them
not_ok %>% arrange(id) %>% select(id:tool_id) %>% head(10)
```

Jobs run per month.

```{r}
# Create new column with year-month
jobs <- jobs %>% 
  mutate(month=year_month(create_time))

# Get number of jobs run per year-month
jobs %>%
  group_by(month) %>% 
  summarize(n=n())
```

```{r include=FALSE}
# ggplot(jobs, aes(x=create_time %>% date)) +
#   geom_histogram(bins=20) +
#   scale_x_date(date_breaks="3 months") +
#   labs(x="date", y="jobs")
# 
# ggplot(jobs, aes(x=create_time %>% ymd_hms)) +
#   geom_density()
```

```{r}
ggplot(jobs, aes(x=month)) + 
  geom_bar(alpha=0.8) +
  labs(x="date", y="number of jobs")
```

## Tool analysis

Which tools are used most often?

```{r}
# Using all jobs across all dates
all_tool_ids <- all_jobs$tool_id
tool_data <- all_tool_ids %>% str_split("/")
tool_names <- sapply(tool_data, function(x) {ifelse(length(x) == 1, x, x[5])})

# Top 20 used tools
tool_names %>% table %>% data.frame %>% setNames(c("tool", "freq")) %>%
  arrange(desc(freq)) %>% head(20)
```

Which reference genomes are frequently used?
```{r}
# For bwa?
bwa_jobs <- all_jobs %>% mutate(tool_name=tool_names) %>% 
  filter(str_detect(tool_name, "bwa"))
bwa_jobs$command %>% 
  str_extract("/mnt/galaxyIndices/[A-Za-z0-9/.-_]+") %>%
  str_split("/") %>% sapply(function(x) x[length(x)]) %>%
  table() %>% data.frame() %>% setNames(c("reference", "freq")) %>%
  arrange(desc(freq))

# For tophat?
tophat_jobs <- all_jobs %>% mutate(tool_name=tool_names) %>% 
  filter(str_detect(tool_name, "tophat"))
tophat_jobs$command %>% 
  str_extract("/mnt/galaxyIndices/[A-Za-z0-9/.-_]+") %>%
  str_split("/") %>% sapply(function(x) x[length(x)]) %>%
  table() %>% data.frame() %>% setNames(c("reference", "freq")) %>%
  arrange(desc(freq))
```

-----

# Session Info

```{r}
devtools::session_info()
```

